pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single-cell RNA-seq preprocessing tools

### 2020-05-25 draft.tex
fixed latex errors

The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of concrete recommendations about analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).

@@

\section*{Background}

Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization. This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}. As data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation, it is critical to evaluate the available tools.

A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\cite{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}). However these recommendations need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \cite{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \cite{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \cite{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}. It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including run time) on the fly. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties. This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 4 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 5), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 6). We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 7-9). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

We next investigated the impact of filtering according to various criteria (see Methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 10). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and SCnorm \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A and Supplementary Figure 12). Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}. Surprisingly, however, regressing out the covariates tended to increase the association with it.

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. An exception was \textit{scVI} (Supplementary Figure 13), which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy. For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{sctransform} (more details in the Methods section). Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 14). We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 15), at little apparent cost on computation time (Supplementary Figure 16).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 17A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 18-19). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 18-20). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 18-19). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 20).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 13). 
Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 14 and 21). This tended to affect the accuracy of the clustering (Supplementary Figure 22), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an 'Elbow' (inflexion point) of the variance explained, to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method (inflexion point in the variance explained by each component), some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.

>>>>

The massive growth of single-cell RNA-sequencing (scRNAseq) and the methods for its analysis still lack sufficient and up-to-date benchmarks that could guide analytical choices. Numerous benchmark studies already exist and cover most of scRNAseq processing and analytical methods but only a few give advice on a comprehensive pipeline. Moreover, current studies often focused on isolated steps of the process and do not address the impact of a tool on both the intermediate and the final steps of the analysis. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics. We apply it to the benchmark of scRNAseq analysis pipelines using simulated and real datasets with known cell identities, covering common methods of filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering. We evaluate the choice of these tools with multi-purpose metrics to assess their ability to reveal cell population structure and lead to efficient clustering. On the basis of our systematic evaluations of analysis pipelines, we make a number of practical recommendations about current analysis choices and for a comprehensive pipeline. The evaluation framework that we developed, \textit{pipeComp} (\url{https://github.com/plger/pipeComp}), has been implemented so as to easily integrate any other step, tool, or evaluation metric allowing extensible benchmarks and easy applications to other fields of research in Bioinformatics, as we demonstrate through a study of the impact of removal of unwanted variation on differential expression analysis.

@@

\section*{Background}
Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization. This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}. As data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation, it is critical to evaluate the available tools.

A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}). However these recommendations need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of a scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}. It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Supplementary Figure 1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one {\color{red}(Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).} Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties. This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 5 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 6), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 7). We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

{\color{red}We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel})}. An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{kuhnHungarianMethod1955} and monitored the mean F1 score {\color{red}across subpopulations. A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. In contrast, too lenient filtering (or lack thereof) can lead to misclassified cells, as was the case for the Zhengmix4 datasets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (\textit{default} - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). PCA-based selection did not appear superior to feature-wise filters. We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B). Of note, doublet removal in conjunction with filtering tended to improve accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g. FACS-sorted datasets). Indeed, the datasets where doublet removal did not have a clear positive impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our \textit{default} filtering (see Methods) or a similar approach.}

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{sctransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 14). Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{sctransform}. Surprisingly, however, regressing out the covariates tended to increase the association with it.

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations{\color{red}, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines). Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal or poorer performance on MI, ARI and ARI at true number of clusters, with a strongest negative impact among the Smart-seq datasets (Koh and Kumar). Regressing out covariates using \textit{sctransform} instead, tended to have a much milder effect, possibly owing to its regularization. In general, \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).}

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 16). We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 17), at little apparent cost on computation time (Supplementary Figure 18).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated ({\color{red}Pearson correlation >0.9,} Supplementary Figure 19A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 20-21). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 20-22). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 20-21). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also {\color{red}tended to lead} to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). {\color{red}Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary} Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. {\color{red}Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).}

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 16 and 24). This tended to affect the accuracy of the clustering (Supplementary Figure 25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an {\color{red}inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms.} We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.
